---
title: "Chapter 3 Vectors"
author: "ADS"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output: html_document
---


# Basic definitions

Vectors are the key group of data types in R. There is a basic division between atomic vectors
and lists:

- **atomic vector**: all elements have the same type
- **list**: elements can have different types

`NULL` is not a vector, but is often used as a generic zero-length vector.

Vectors can have **attributes**: a named list of metadata. Two of these attributes are widely used:

- **dimension**: by adding dimensions, vectors can be structured as matrices
- **class**: this attribute is central to the S3 object system.

# Section 3.2 Atomic vectors

Atomic vectors include the following types:

1. logical
2. integer
3. double
4. character (string)
5. complex
6. raw

The first four are common. The types `integer` and `double` are both `numeric`.

The types `complex` and `raw` are used less often. Complex numbers are stored in vectors of type
`complex`, and raw bytes (binary data) are stored in vectors of type `raw`.

## "Scalar" literals

R actually implements single values as vectors of length 1.

```{r}
x <- 1
x[1]

# or even
TRUE[1]
# this one only works in the console
# 1[1]

```

There are literals to represent a vector of length 1 for each type.

- logical: `TRUE` or `FALSE`, or `T` or `F`
- double: decimal `0.1231`, scientific `1.23e4` or hex `0xCAFE`, 
  plus three special values `Inf`, `-Inf`, `NaN`
- integer: decimal `1231L`, scientific `1.23e4L` or hex `0xCAFEL`
- string: double or single quotation marks `"Hi!"`, `'Bye!'`, but double quotes are preferred

```{r}
a <- TRUE
b <- 3.14
c <- 3.14e-1
d <- 0xBEEF
e <- "Hi!"
f <- 'Bye!'

a
b
c
d
e
f
```

String content often requires escaping: `?Quotes` gives the details.

Remember here that non-syntactic names can be "escaped" by quoting with backticks: `.2way`.
