---
title: 'Chapter 2: names and values'
author: "ADS"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output: html_document
---

```{r setup}
library(lobstr)
```

# Section 2.2: Binding basics

Create an object and bind it to a name:

```{r}
x <- c(1, 2, 3)
```

Bind the object to another name:

```{r}
y <- x
```

Only one object exists at this stage: check this by inspecting the
object's address in memory:

```{r}
obj_addr(x)
obj_addr(y)
```


Names need to avoid reserved words (see `?Reserved`), follow the naming rules 
(see `?make.names`) or be surrounded with backticks.

**ALARM!** The locale of the system determines what are "letters" and what are
not letters. This has unfortunate side-effects:

> "That means that the syntax of R code cab actually differ from computer to
> computer, and that it's possible for a file that works on one computer to not
> even parse on another! Avoid this problem by sticking to ASCII characters
> (i.e. A-Z) as much as possible."
> 
> --- Wickham 2019, 20 n. 1.

## 2.2 Exercises

### Question 1

```{r}
a <- 1:10
b <- a
c <- b
d <- 1:10
```

All three of `a`, `b`, and `c` are names bound to the same object; `d` is bound 
to a different object.

### Question 2

```{r}
mean
base::mean
get("mean") # `base::get` "return the value of a named object"
evalq(mean) # `base::evalq` "evaluate an R expression in an environment"
match.fun("mean")
```

All these expressions point to the same underlying object.

### Question 3

The default conversion to syntactic names is problematic because functions like 
read.csv() (see `?read.csv`) silently change the column names in the source 
data. Suppress this with `check.names = FALSE`.

### Question 4

Rules for `make.names` (see `?make.names`):

> The character "X" is prepended if necessary. All invalid characters are
> translated to ".". A missing value is translated to "NA". Names which match R
> keywords have a dot appended to them. Duplicated values are altered by
> `make.unique`.
> 
> --- [Package documentation `base` version 4.0.5]

### Question 5

`.123e1` is not a syntactic name because syntactic names start with a letter or
a dot *not followed by a number*.

# Section 2.3: copy-on-modify

R has **copy-on-modify** behaviour for the bindings of names to values. You can bind multiple names
to the same value, but when you modify the value of one of these bindings, a new object is created
and bound to that name. In general, R objects are immutable, but see section 2.5 for instances of
modify-in-place behaviour.

```{r}
x <- c(1, 2, 3)
y <- x
# same object
obj_addr(x)
obj_addr(y)
# modify the contents of y
y[[3]]
y[[3]] <- 4
# different contents and different object
x
y
obj_addr(x)
obj_addr(y)
```

To explore copy-on-modify, you need to use a terminal or RMarkdown instead of the RStudio console.
RStudio's environment pane keeps references to objects in order to print information about them,
and this confuses things.

```{r}
x
# turn on memory tracing for the object bound to the name x
cat(tracemem(x), "\n")
y <- x
# when y is first modified, a new copy is made
y[[3]] <- 4L
# but not after that, due to modify-in-place optimisation (only one name bound to the object)
y[[3]] <- 5L
# turn off memory tracing
untracemem(x)
```

This behaviour also applies to function calls. To demonstrate this, define an identify function
with its own internal name-value bindings, and use the memory tracing technique.

```{r}
# identity function with internal binding
f <- function(a) {
  a
}

x <- c(1, 2, 3)
cat(tracemem(x), "\n")

z <- f(x)
# z and x point to the same object: no copy
# but a function call involving a modification will generate a copy
g <- function(b) {
  b[[1]] <- 100
}

y <- g(x)

untracemem(x)

```

The same behaviour applies to lists, but lists store **references** to values rather than the
values themselves.

The copy-on-modify behaviour for lists is to make a **shallow copy** (since R 3.1.0): references to
the changed values are updated when the copy is made, but the references to the unchanged values
remain the same.

The function `lobstr::ref()` exposes the relevant memory addresses.

```{r}

l1 <- list(1, 2, 3)
l2 <- l1
l2[[3]] <- 4

ref(l1, l2)
```

Data frames are lists of vectors. Note that modifying a column changes only that column, but that
modifying a row changes all the columns: in the context of copy-on-modify, this makes the first
operation much more efficient than the latter.

Character vectors are also references under the hood, specifically to a global string pool, in order
to make memory usage more efficient.

## 2.3 Exercises

### Question 1

`tracemem(1:10)` is not useful because there is no binding between a name and a value, just the
value itself, and so there is therefore nothing to track (the binding of names to objects).

### Question 2

A copy is made when setting `x[[3]]` to a double, since up to this point `x` is an integer vector.

### Question 3

`a` is bound to an object with the value `1:10`; `b` is bound to a list with two references to `a`;
`c` is bound to a list with references to `b`, to `a`, and to another different object with the
value `1:10`.

### Question 4

```
x <- list(1:10)
x[[2]] <- x
ref(x)
#> o [1:0xd0c8a58] <list>
#> +-[2:0xc709b30] <int>
#> \-o [3:0xd139fd0] <list>
#>   \-[2:0xc709b30]
```

Setting `x[[2]] <- x` makes a copy of the list and inserts the copy into the original list, but
the same object with the value `1:10` is referenced at both levels.

# 2.4 Object size

Use `lobstr::obj_size()` for a measure of the memory that an object takes. Avoid the similar
function `utils::object.size()`.

Lists and character vectors take up much less memory than might be expected due to the use of 
references.

Since R 3.5 the ALTREP optimisation leads to compact storage of certain types of vectors, e.g. 
numeric vectors involving ranges.

```{r}
obj_size(1:3)
obj_size(1:1e3)
obj_size(1:1e6)
obj_size(1:1e9)
# all 680 B
```

## 2.4 Exercises

### Question 1

```{r}
y <- rep(list(runif(1e4)), 100)

object.size(y)
#> 8005648 bytes
obj_size(y)
#> 80,896 B
```

Since `object.size` doesn't account for shared objects, the hundred-fold repetition of the same
object `list(runif(1e4))` is not taken into account.

```{r}
x <- runif(1e4)
obj_size(x)
#> 80,048 B

```

### Question 2

```{r}
funs <- list(mean, sd, var)
obj_size(funs)
#> 17,608 B
```


The list comprises three references to functions in base; these are always there, as part of R
itself, and there may be underlying internal code that is not represented in the surface R object.

### Question 3

```{r}
a <- runif(1e6)
obj_size(a)

b <- list(a, a)
obj_size(b)
obj_size(a, b)

b[[1]][[1]] <- 10
obj_size(b)
obj_size(a, b)

b[[2]][[1]] <- 10
obj_size(b)
obj_size(a, b)
```

`a` will be a certain size, given the amount of numbers it contains. `b` will be slightly larger,
with two references to the same object.

`a` and `b` together will be the same size as `b`, given that `b` contains `a`.

Once the first element of `b` is modified, then the overall size of `b` will double: there are now
two of the structure initially represented as `a`. `a` and `b` together will be the same size
as `b`, given that `b` still contains `a`.

Once the second element of `b` is modified, the overall size of `b` remains as before: two of the
structure initially represented as `a`. But `a` and `b` together will now be three times as large
as `a`: the two structures in `b` are no longer the same as `a` and there are now three copies of 
the structure initially represented as `a`.

# 2.5 Modify-in-place

There are two exceptions to the copy-on-modify behaviour:

1. A object with a single binding is modified-in-place, as an optimisation (no extra allocation).
2. Environments are always modified-in-place, as an exception to the rule.

Things get a bit more complicated. 
