---
title: 'Chapter 2: names and values'
author: "ADS"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output: html_document
---

```{r setup}
library(lobstr)
```

# Section 2.2: Binding basics

Create an object and bind it to a name:

```{r}
x <- c(1, 2, 3)
```

Bind the object to another name:

```{r}
y <- x
```

Only one object exists at this stage: check this by inspecting the
object's address in memory:

```{r}
obj_addr(x)
obj_addr(y)
```


Names need to avoid reserved words (see `?Reserved`), follow the naming rules 
(see `?make.names`) or be surrounded with backticks.

Note that the locale of the system determines what are "letters" and what are
not letters. This has unfortunate side-effects:

> "That means that the syntax of R code cab actually differ from computer to
> computer, and that it's possible for a file that works on one computer to not
> even parse on another! Avoid this problem by sticking to ASCII characters
> (i.e. A-Z) as much as possible."
> 
> --- Wickham 2019, 20 n. 1.

## 2.2.2 Exercises

### Question 1

```{r}
a <- 1:10
b <- a
c <- b
d <- 1:10
```

All three of `a`, `b`, and `c` are names bound to the same object; `d` is bound 
to a different object.

### Question 2

```{r}
mean
base::mean
get("mean") # `base::get` "return the value of a named object"
evalq(mean) # `base::evalq` "evaluate an R expression in an environment"
match.fun("mean")
```

All these expressions point to the same underlying object.

### Question 3

The default conversion to syntactic names is problematic because functions like 
read.csv() (see `?read.csv`) silently change the column names in the source 
data. Suppress this with `check.names = FALSE`.

### Question 4

Rules for `make.names` (see `?make.names`):

> The character "X" is prepended if necessary. All invalid characters are
> translated to ".". A missing value is translated to "NA". Names which match R
> keywords have a dot appended to them. Duplicated values are altered by
> `make.unique`.
> 
> --- [Package documentation `base` version 4.0.5]

### Question 5

`.123e1` is not a syntactic name because syntactic names start with a letter or
a dot *not followed by a number*.

# Section 2.3: copy-on-modify


